Spatial Reasoning and Data Displays

### 2015-01-07 Paper/Draft1/TestingVisualAptitude.Rnw
comments throughout, but just minor ones

\comments{This is going into the right direction, but a few pieces are still missing, 
you need to still say what this paper is doing: we do a study to compare lineup performance with established tests, also go into a bit of discussion of the tests.}
% \comment{Here is a paragraph missing, that connects the existing literature to what we are doing in this paper. 
% In particular, we need to answer the questions: 
% 
% - why are we interested in evaluating which skills are necessary to evaluate graphical displays (this has been implicitly answered by saying that cognitive load was reduced, so we free up resources and make graphics a better tool for more people.
% 
% - what are the limits of the studies that you have cited, ie is there something that we can do better?
% 
% - what advantage do lineups bring to the table?
% }
% 
% 
% \comment{How do perceptual principles apply to statistical graphics (data displays)? - Lineups help us to isolate different parts of displays and allow in-built quantification of results.}
% 
% 
% \comment{Why are lineups variable? - Individual variability is big, we want to investigate which abilities individuals help the most in evaluating. - for this reason we need to motivate why lineups are useful. }
% 
% 
% \comment{We examine 'some demographic characteristics' is a bit lukewarm. I think once the missing paragraph is there, you can sell the rest of the paper a bit more passionately.}

@@

qplot(data=lineup.summary.categorical, x=value, y=lineup, geom="boxplot", fill=I("grey70")) + 
  facet_wrap(~variable, scales="free_x", ncol=4) + xlab("") + 
  ylab("Scaled Lineup Score") + 
  geom_point(aes(x=value, y=lineup), shape=1, size=3) + 
  theme_bw() + 
  theme(plot.margin=unit(c(0,0,0,0), unit="mm"))
@
\begin{figure*}[h!tb]\centering
\includegraphics[width=.9\linewidth]{fig-VisReasoningCategorical-1}
\caption[Visual Aptitude Study Results]{Demographic characteristics of participants compared with lineup score. Categories are ordered by effect size; majoring in a STEM field, calculus completion, hours spent playing video games per week, and sex are all associated with a significant difference in lineup score. }\label{fig:visualaptitudecat}
\end{figure*}

@@

From figure~\ref{fig:lineupblockpca} we see that again, PC1 accounts for general visual aptitude. 
% PC2 emphasizes the relationship between figure classification and lineup task 1 (and to a lesser extent, on lineup task 2). PC3 differentiates the lineup task 3 from the visual search task; other tasks do not impact PC3 to a meaningful degree. PC4 separates lineup test 2 from the paper folding and card rotation tests and emphasizes the overlap with the visual search test. PC5 separates lineup test 2 and the figure classification test from the overlap between the other lineup tests and the visual search test. PC6 and PC7 account for a joint total of less than 10\% of the overall variability, and do not display any distinct patterns in the loadings.
%
PC2 emphasizes the overlapping variation in performance on lineup test 1 and the figure classification test. PC3 emphasizes the additional variation in performance on lineup test 3 and the visual search test, while PC4 emphasizes the extra variability in performance on lineup test 2 and the paper folding test. All three lineups, plus the figure classification, paper folding, and visual search tests contribute to PC5. PC6 and PC7 jointly account for less than 10\% of the variance in the data and do not display any distinct patterns in the loadings. 
%\Sexpr{as.numeric(round(100*importance[6], 1))}\% and \Sexpr{as.numeric(round(100*importance[7], 1))}\% of the variance, respectively; PC6 seems to differentiate the first two lineup tasks from the overlap between the third lineup task and the figure classification task.

@@

\section{Discussion and Conclusions}
Performance on lineups is strongly related to performance on tests of visual ability; however, this relationship is mediated by factors such as mathematical training, hours of video games played per week, and gender.
In addition, many facets of intelligence are highly correlated; those who score higher on general aptitude tests may score higher on tests of visual ability (and may also score higher on lineup tests).
Despite these caveats, we have demonstrated that the general lineup task is most closely related to a classification task, rather than tests of spatial ability.
This is an important verification of a tool that is useful for examining statistical graphics, as it emphasizes the idea that while the testing medium is graphical in nature, the task is in fact a classification task, where the viewer must determine the most important features of each plot and then identify which plot is different.

@@

\section{Lineup Performance and Demographic Characteristics}\label{app:categoricalresults}
Table~\ref{tab:ttest-demographics} provides the results of a sequence of linear models fit to the lineup data. Each row in the table represents a single model, with one predictor variable (a factor with two or more levels). Due to sample size considerations, multiple testing corrections were not performed; in addition, the independent variables are correlated: in our sample, males are more likely to have completed Calculus 1, but are also more likely to spend time playing video games. As such, a model including two or more of the significant predictor variables shows all included variables to be nonsignificant. To better understand the effects of these variables, a larger study would be required. 
<<ttests.categ, echo=F, eval=T, results='asis'>>=
tmp <- dlply(lineup.summary.categorical, .(variable), function(df){
    tmp <- anova(lm(lineup~factor(value, ordered=F), data=df))
    coefs <- as.data.frame(tmp)[1,]
    coefs$level <- gsub("factor(value, ordered = F)", unique(df$variable), rownames(coefs), fixed=T)
    rownames(coefs) <- NULL
    coefs <- coefs[,c(6, 1:5)]
    names(coefs) <- c("Variable", "DF", "Sum.of.Squares", "MeanSq", "F", "p.val")
    coefs
})

@@

\section{Principal Component Analysis of Visuo-Spatial Tests}\label{app:pca}
\subsection{PCA of the Four Cognitive Tests}\label{app:pca4}
%
<<pcasummary4,echo=F, results='asis'>>=
pca2 <- prcomp(ans.summary[,c("card_rot", "fig_class", "folding", "vis_search")], scale=T, retx=T)
print(xtable(summary(pca2), caption="Importance of principal components in an analysis of four tests of spatial ability: figure classification, paper folding, card rotation, and visual search.\\label{tab:PCAvariance4}",digits=2), sanitize.text.function=sanitize, sanitize.rownames.function=sanitize, sanitize.colnames.function=sanitize, size="footnotesize", table.placement="htb", caption.placement="top")
@
Table~\ref{tab:PCAvariance4} contains the importance of each resultant PC and the proportion of the variance each PC represents.

@@

<<biplot-pca4, fig.cap="", fig.width=4.5, fig.height=4.5, warning=F, message=F, out.width='.7\\linewidth', comment="|", fig.cap="Plots of principal components 1-4 with observations. PCA was performed on the four cognitive tests used to understand the cognitive demands of the lineup protocol.  \\label{fig:biplots4}", echo=F, fig.env="figure">>=
par(mar=c(4,4,.1,.1),cex.lab=.95,cex.axis=.9,mgp=c(2,.7,0),tcl=-.3)
biplot(pca2, choices=1:2, pc.biplot=T, cex=c(.5, 1), xlim=c(-2.5,2.5), ylim=c(-2.5,2.5))
biplot(pca2, choices=3:4, pc.biplot=T, cex=c(.5, 1), xlim=c(-3,3), ylim=c(-2.5,2.5))
@

\subsection{PCA of Cognitive Tests and Lineups}\label{app:pca5}
PC1 is essentially an average across all tests representing a general ``visual intelligence" factor.
Biplots of the remaining principal components are shown in figure~\ref{fig:biplots5}.

>>>>

\comment{This is going into the right direction, but a few pieces are still missing, 
you need to still say what this paper is doing: we do a study to compare lineup performance with established tests, also go into a bit of discussion of the tests.}
% \comment{Here is a paragraph missing, that connects the existing literature to what we are doing in this paper. 
% In particular, we need to answer the questions: 
% 
% - why are we interested in evaluating which skills are necessary to evaluate graphical displays (this has been implicitly answered by saying that cognitive load was reduced, so we free up resources and make graphics a better tool for more people.
% 
% - what are the limits of the studies that you have cited, ie is there something that we can do better?
% 
% - what advantage do lineups bring to the table?
% }
% 
% 
% \comment{How do perceptual principles apply to statistical graphics (data displays)? - Lineups help us to isolate different parts of displays and allow in-built quantification of results.}
% 
% 
% \comment{Why are lineups variable? - Individual variability is big, we want to investigate which abilities individuals help the most in evaluating. - for this reason we need to motivate why lineups are useful. }
% 
% 
% \comment{We examine 'some demographic characteristics' is a bit lukewarm. I think once the missing paragraph is there, you can sell the rest of the paper a bit more passionately.}

@@

qplot(data=lineup.summary.categorical, x=value, y=lineup, geom="boxplot", fill=I("grey")) + 
  facet_wrap(~variable, scales="free_x", ncol=4) + xlab("") + 
  ylab("Scaled Lineup Score") + 
  geom_point(aes(x=value, y=lineup), shape=1, size=3) + 
  theme_bw() + 
  theme(plot.margin=unit(c(0,0,0,0), unit="mm"))
@
\begin{figure*}[h!tb]\centering
\includegraphics[width=.9\linewidth]{fig-VisReasoningCategorical-1}
\caption[Visual Aptitude Study Results]{Demographic characteristics of participants compared with lineup score. Categories are ordered by effect size; majoring in a STEM field, calculus completion, hours spent playing video games per week, and sex are all associated with a significant difference in lineup score. }\label{fig:visualaptitudecat}
\end{figure*}

@@

\comment{First: what is shown in figure 8? THis is a non traditional plot, so you need to explain it first.}
From figure~\ref{fig:lineupblockpca} we see that again, PC1 accounts for general visual aptitude. 
% PC2 emphasizes the relationship between figure classification and lineup task 1 (and to a lesser extent, on lineup task 2). PC3 differentiates the lineup task 3 from the visual search task; other tasks do not impact PC3 to a meaningful degree. PC4 separates lineup test 2 from the paper folding and card rotation tests and emphasizes the overlap with the visual search test. PC5 separates lineup test 2 and the figure classification test from the overlap between the other lineup tests and the visual search test. PC6 and PC7 account for a joint total of less than 10\% of the overall variability, and do not display any distinct patterns in the loadings.
%
PC2 emphasizes the overlapping variation in performance on lineup test 1 and the figure classification test. PC3 emphasizes the additional variation in performance on lineup test 3 and the visual search test, while PC4 emphasizes the extra variability in performance on lineup test 2 and the paper folding test. All three lineups, plus the figure classification, paper folding, and visual search tests contribute to PC5. PC6 and PC7 jointly account for less than 10\% of the variance in the data and do not display any distinct patterns in the loadings. 
%\Sexpr{as.numeric(round(100*importance[6], 1))}\% and \Sexpr{as.numeric(round(100*importance[7], 1))}\% of the variance, respectively; PC6 seems to differentiate the first two lineup tasks from the overlap between the third lineup task and the figure classification task.

@@

\section{Discussion and Conclusions}
\comment{These read like the conclusions from earlier - video games aren't that prominently displayed anymore unless you are referring to figure 5. But if so, I would be cautious about summarising STEM major as 'mathematical training' - use the same names for the variables as in figure 5 and make sure to keep the order, i.e. gender is more impotant than video game hours. also make sure, that to mention that this is on a univariate level and doesn't take the correlation between the variables into account. }

Performance on lineups is strongly related to performance on tests of visual ability; however, this relationship is mediated by factors such as mathematical training, hours of video games played per week, and gender.
In addition, many facets of intelligence are highly correlated; those who score higher on general aptitude tests may score higher on tests of visual ability (and may also score higher on lineup tests).
Despite these caveats, we have demonstrated that the general lineup task is most closely related to a classification task, rather than tests of spatial ability.
This is an important verification of a tool that is useful for examining statistical graphics, as it emphasizes the idea that while the testing medium is graphical in nature, the task is in fact a classification task, where the viewer must determine the most important features of each plot and then identify which plot is different.

@@

\section{Lineup Performance and Demographic Characteristics}\label{app:categoricalresults}
\comment{'Results' of lineup score - te caption in the table is too generic. give more details. }
Table~\ref{tab:ttest-demographics} provides the results of a sequence of linear models fit to the lineup data. Each row in the table represents a single model, with one predictor variable (a factor with two or more levels). Due to sample size considerations, multiple testing corrections were not performed; in addition, the independent variables are correlated: in our sample, males are more likely to have completed Calculus 1, but are also more likely to spend time playing video games. As such, a model including two or more of the significant predictor variables shows all included variables to be nonsignificant. To better understand the effects of these variables, a larger study would be required. 
<<ttests.categ, echo=F, eval=T, results='asis'>>=
tmp <- dlply(lineup.summary.categorical, .(variable), function(df){
    tmp <- anova(lm(lineup~factor(value, ordered=F), data=df))
    coefs <- as.data.frame(tmp)[1,]
    coefs$level <- gsub("factor(value, ordered = F)", unique(df$variable), rownames(coefs), fixed=T)
    rownames(coefs) <- NULL
    coefs <- coefs[,c(6, 1:5)]
    names(coefs) <- c("Variable", "DF", "Sum.of.Squares", "MeanSq", "F", "p.val")
    coefs
})

@@

\section{Principal Component Analysis of Visuo-Spatial Tests}\label{app:pca}
\subsection{PCA of the Four Cognitive Tests}\label{app:pca4}
\comment{'Importance of principal components {\it in an analysis} of four tests' -- be more precise: you mean the PCA. Give some details as to how importance is measured. }
%
<<pcasummary4,echo=F, results='asis'>>=
pca2 <- prcomp(ans.summary[,c("card_rot", "fig_class", "folding", "vis_search")], scale=T, retx=T)
print(xtable(summary(pca2), caption="Importance of principal components in an analysis of four tests of spatial ability: figure classification, paper folding, card rotation, and visual search.\\label{tab:PCAvariance4}",digits=2), sanitize.text.function=sanitize, sanitize.rownames.function=sanitize, sanitize.colnames.function=sanitize, size="footnotesize", table.placement="htb", caption.placement="top")
@
Table~\ref{tab:PCAvariance4} contains the importance of each resultant PC and the proportion of the variance each PC represents.

@@

<<biplot-pca4, fig.cap="", fig.width=4.5, fig.height=4.5, warning=F, message=F, out.width='.7\\linewidth', comment="|", fig.cap="Plots of principal components 1-4 with observations. PCA was performed on the four cognitive tests used to understand the cognitive demands of the lineup protocol.  \\label{fig:biplots4}", echo=F, fig.env="figure">>=
par(mar=c(4,4,.1,.1),cex.lab=.95,cex.axis=.9,mgp=c(2,.7,0),tcl=-.3)
biplot(pca2, choices=1:2, pc.biplot=T, cex=c(.5, 1), xlim=c(-2.5,2.5), ylim=c(-2.5,2.5))
biplot(pca2, choices=3:4, pc.biplot=T, cex=c(.5, 1), xlim=c(-3,3), ylim=c(-2.5,2.5))
@
\comment{figure 9 and 10 are Biplots, say so in the caption - in the discussion of the figures and the tables in the text, please think again of adding the sentence summarising what is shown. }
\subsection{PCA of Cognitive Tests and Lineups}\label{app:pca5}
PC1 is essentially an average across all tests representing a general ``visual intelligence" factor.
Biplots of the remaining principal components are shown in figure~\ref{fig:biplots5}.
